---
title: "MDSS Datathon"
author: "Arindam Baruah, Aishwarya Anil Kumar"
format: html
editor: visual
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(ggplot2)
library(visdat)
library(janitor)
library(plotly)
library(ggthemes)
library(scales)
library(RColorBrewer)
library(treemap)
library(ozmaps)
library(sf)
```
# Key questions

1. Which states are the leading waste generators in Australia ?



```{r}
df <- read_csv("data-raw/data_waste.csv")
head(df)
```



Let us clean the names of the variables

```{r}
df_waste <- clean_names(df)


df_waste[c("start_year","end_year")] <- (str_split_fixed(df_waste$year,"-",2))
df_waste$start_year <- as.numeric(df_waste$start_year)
df_waste$end_year <- as.numeric(df_waste$end_year)

df_waste<- df_waste%>% select(-year) 

head(df_waste)
```
# Waste per state


```{r}
df_waste_state <- df_waste %>% filter(stream == "Total") %>%  group_by(jurisdiction,category) %>%
  summarise(Total_tonnes = sum(tonnes)) %>% filter(jurisdiction !="Australia" & category != "Total")

df_waste_grouped <- df_waste_state %>% group_by(jurisdiction) %>% 
  summarise(Total_tonnes = sum(Total_tonnes))
df_waste_grouped <- df_waste_grouped %>% 
  mutate(jurisdiction = case_match(jurisdiction, "ACT"~"Australian Capital Territory",
                                                                          "NSW"~"New South Wales",
                                                                          "Vic" ~ "Victoria",
                                                                          "Qld" ~ "Queensland",
                                                                          "SA" ~ "South Australia",
                                                                          "WA" ~ "Western Australia",
                                                                          "NT" ~ "Northern Territory",
                                                                          "Tas" ~ "Tasmania"))


sf_oz <- ozmap("states")
sf_oz <- left_join(sf_oz,df_waste_grouped,by=c("NAME"="jurisdiction"))

label_data <- data.frame(
  state = c("NSW", "VIC", "QLD", "SA", "WA", "TAS"),
  lon = c(146, 144, 146, 135, 122, 147),
  lat = c(-33.85, -37.8, -19.25, -30.0, -25.0, -42.0)
)

pl <- ggplot(data = sf_oz, aes(fill = Total_tonnes/10^6)) + geom_sf() + scale_fill_gradient(low ="green", high = "red")+
  theme_classic() + labs(fill = "Total generated waste \n (Megatonnes)",title = "Waste generated in each state of Australia",
                         subtitle = "2006-2021")
pl


df_waste_grouped <- df_waste_grouped %>% mutate(Proportion = 100* Total_tonnes/sum(Total_tonnes))
```
We can observe that:

1. WA leads the way in waste generation followed by NSW and QLD.
2. While this maybe counter intuitive as WA has a considerably smaller population (2.67 million) when compared to NSW (8.16 million) and QLD (5.185 million), however, due to the presence of heavy industries such as mining, petroleum and chemicals, a significant amount of waste is generated in WA.
3. The most common form of waste generated from each state are observed to be ash and rubble.



```{r}

pl1 <- treemap(df_waste_state,
            index=c("category","jurisdiction"),
            vSize="Total_tonnes",
        type = "index",title = "Types of waste in each state of Australia (2006-2021)") 

```




# Hazardous waste

```{r}
df_haz <- df_waste %>% filter(category == "Hazardous wastes")
df_haz_tot <- df_haz %>% filter(stream == "Total")
df_haz_total <- df_haz_tot %>% group_by(start_year,jurisdiction,category) %>% 
  summarise(Tonnes = sum(tonnes)) %>% 
  ungroup() 



```

```{r}
pl1 <- ggplot(data = df_haz_total,aes(x = start_year,y = Tonnes/10^6)) +
  geom_line() + geom_point(color='red') + theme(axis.text.x = element_text(angle = 45,size=5)) + facet_wrap(~jurisdiction) + labs(x = "Year",y = "Mega tonnes",title = "Hazardous waste generated in each state",subtitle = "2006-2021") + theme_classic()
pl1
```

1.  NSW accounts for the highest hazardous waste in the whole of Australia with close to 8.5 MT. The hazardous wastes are on a rising trend.
2.  QLD is observed to record constant generation of hazardous waste each year with approximate 6 MT.
3.  VIC is observed to record a rising trend of hazardous waste in recent years with close to 5.75 MT in 2021.

## How is the Hazardous waste dealt with in each state ?

```{r}

haz_jurisdictions <- c("NSW","Vic","Qld")
df_haz_subset <- df_haz_tot %>% filter(jurisdiction %in% haz_jurisdictions)

df_haz_man <- df_haz_subset %>% group_by(jurisdiction,management) %>% 
  summarise(Tonnes = sum(tonnes)) %>% ungroup()


pl2 <- ggplot(data = df_haz_man,aes(x = reorder(management,Tonnes),y = Tonnes/10^6)) + geom_col(fill = "lightgray",color = "black") + 
  facet_wrap(~jurisdiction) + coord_flip() + geom_col(data = ~filter(.x, management == "Landfill"),fill = "darkred") + geom_label(data = ~filter(.x, management == "Landfill"),aes(label = round(Tonnes/10^6,1),hjust=1)) + geom_col(data = ~filter(.x, management == "Energy from waste facility"),fill = "green") + geom_label(data = ~filter(.x, management == "Landfill"),aes(label = round(Tonnes/10^6,1),hjust=1)) + geom_label(data = ~filter(.x, management == "Energy from waste facility"),aes(label = round(Tonnes/10^6,1),hjust=-0.5)) + 
  labs(x = "Waste management technique",y = "Mega tonnes of waste",title ="Hazardous waste management in the top 3 \n waste producing states",subtitle = "2006-2021") + theme_economist() 
pl2
```

We observe that the:

1.  In all the three states (NSW,Qld,Vic), majority of the hazardous waste are managed through disposing in landfills. While this may be a temporary solution, the implications can be quite devastating **as such waste normally take years to decompose and start polluting the water ways as well as pottable water resources.**

2.  While there have been efforts to utilise hazardous waste to produce energy, it is still a long way behind, with more concerted efforts required by each state to curb disposal of hazardous waste into landfills and obtain clean energy.

3.  While **VIC is observed to generate lesser hazardous waste than NSW and QLD, they're also leading with the highest energy production from waste facilities** with approximately 1.6 Mega tonnes of waste utilised.

# What is the constitution of these Hazardous wastes ?

```{r}

hazardous_waste_summary <- df_haz %>%
  group_by(type) %>%
  filter(jurisdiction %in% haz_jurisdictions) %>%
  summarise(Total_Tonnage = sum(tonnes))

hazardous_waste_summary <- hazardous_waste_summary %>%
  filter(!is.na(type) & !is.na(Total_Tonnage))


ggplot(hazardous_waste_summary, aes(x = reorder(type, Total_Tonnage/10^6), y = Total_Tonnage/10^6)) +

  # Lollipop stems
  geom_segment(aes(y = 0, 
                   yend = Total_Tonnage/10^6, 
                   x = reorder(type, Total_Tonnage/10^6),
                   xend = reorder(type, Total_Tonnage/10^6)), 
               color = "black", size = 1) +
  
  # Lollipop heads
  geom_point(aes(fill = type), size = 6, shape = 21, color = "white") +

  # Y-axis scale formatting
  scale_y_continuous(labels = scales::label_comma()) +
  
  # Chart title and axes labels
  labs(title = "Composition of Hazardous Wastes", 
       x = "Type of Waste", 
       y = "Mega tonnes of waste") +

  # Classic theme and customization
  theme_classic() +
  coord_flip() +
  theme(legend.position = "none", 
        axis.text.y = element_text(size = 8, color = "black"),
        axis.title = element_text(face = "bold", size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        panel.background = element_rect(fill = "cornsilk"),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(linewidth = 0.1, color = "black"))


```
```{r}

haz_jurisdictions <- c("NSW","Vic","Qld")
df_haz_subset <- df_haz_tot %>% filter(jurisdiction %in% haz_jurisdictions)

df_state_total <- df_haz_tot %>% filter(jurisdiction %in% haz_jurisdictions) %>%
  group_by(jurisdiction) %>%
  mutate(total_waste_per_state = sum(tonnes)) %>%
    select(jurisdiction, tonnes, management, total_waste_per_state) %>%
  ungroup()

df_haz_man <- df_state_total %>% 
  filter(jurisdiction %in% haz_jurisdictions) %>%
  group_by(jurisdiction, management) %>%
  summarise(Tonnes = sum(tonnes)) %>%
  left_join(df_state_total %>% 
              group_by(jurisdiction) %>%
              summarise(total_waste_per_state = sum(tonnes)), by = "jurisdiction") %>%
  mutate(percent_by_management = round(Tonnes / total_waste_per_state * 100, 2)) %>%
  select(jurisdiction, management, total_waste_per_state, Tonnes, percent_by_management)

upper_limit <- ceiling(max(df_haz_man$percent_by_management, na.rm = TRUE) / 10) * 10

pl2 <- ggplot(data = df_haz_man, aes(x = reorder(management, percent_by_management), y = percent_by_management)) + 
  geom_col(fill = "lightgray", color = "black") + 
  facet_wrap(~jurisdiction) + 
  coord_flip() + 
  geom_col(data = ~filter(.x, management == "Landfill"), fill = "darkred") + 
  geom_label(data = ~filter(.x, management == "Landfill"), aes(label = round(percent_by_management, 1), hjust=1)) + 
  geom_col(data = ~filter(.x, management == "Energy from waste facility"), fill = "green") + 
  geom_label(data = ~filter(.x, management == "Energy from waste facility"), aes(label = round(percent_by_management, 1), hjust=-0.5)) + 
  labs(x = "Waste management technique", y = "Total percentage of waste", title ="Hazardous waste management in the top 3 \n waste producing states", subtitle = "2006-2021") + 
  theme_economist() + 
  scale_y_continuous(breaks=seq(0, upper_limit, 10), labels = scales::label_number(scale = 1)) + theme(axis.text.x = element_text(size =6))

pl2
```

```{r}


df_haz_source <-df_haz %>% filter(jurisdiction %in% haz_jurisdictions,stream != "Total")

df_haz_source_grouped <- df_haz_source %>%  group_by(jurisdiction,type,stream) %>% summarise(Total_tonnes =  sum(tonnes))

df_haz_source_grouped <- df_haz_source_grouped %>% mutate(stream = case_match(stream,
                                                                              "C&D"~"Construction and Demolition",
                                                                              "C&I" ~ "Commercial and Industrial"))

pl3 <- ggplot(data = df_haz_source_grouped %>% drop_na(),
              aes(
                x = reorder(type, Total_tonnes),
                y = Total_tonnes / 10 ^ 6,
                fill = stream
              )) + geom_col() + coord_flip() + labs(x = "Type of hazardous waste", y = "Mega tonnes of waste", fill = "Source of waste") + theme_classic() + theme(
                panel.background = element_rect(fill = "cornsilk"),
                axis.text.x = element_text(size = 6, face = "bold"),
                axis.text.y = element_text(size = 6 , face = 'bold'),
                legend.text = element_text(size = 6),
                legend.position = "bottom"
              ) + facet_wrap( ~jurisdiction) + ggtitle("Source of each form of hazardous waste", subtitle = "2006-2021") + scale_fill_brewer(palette = "Set2") + geom_col(data = ~filter(.x, type == "Asbestos (N220)"),color = "red") 

pl3
```

Asbestos has been observed to be clearly the differentiating factor for NSW hazardous waste generation when compared to QLD and VIC with close to equal contribution from the commercial & industrial and construction & demolition.


How is Asbestos handled ?

```{r}

df_haz_source_fate <- df_haz_source %>% filter(type == "Asbestos (N220)") %>% 
  group_by(jurisdiction,stream,fate) %>% 
  summarise(Total_tonnes = sum(tonnes))

df_haz_source_fate <- df_haz_source_fate %>% mutate(stream = case_match(stream,
                                                                              "C&D"~"Construction and Demolition",
                                                                              "C&I" ~ "Commercial and Industrial"))

pl4 <- ggplot(data = df_haz_source_fate,aes(x = fate,y = Total_tonnes/10^6,fill = stream)) + 
  geom_col(color = "black") + 
  facet_wrap(~jurisdiction) + theme_minimal() + theme(panel.background = element_rect(fill = "cornsilk"),axis.text.x = element_text(angle =40,vjust = 0.8,size=6.5,face = "bold")) + labs(x ="Type of waste management",y = "Mega tonnes of waste",fill = "Source of waste") + ggtitle("Waste management of Asbestos in \n top Hazardous waste producing states",subtitle = "2006-2021") + scale_fill_brewer(palette = "Reds")
pl4
```
We can observe that Asbestos does indeed pose a significant threat to the environment with no known techniques to either recycle or generate energy out of it. As a result, nearly all the generated waste out of Asbestos needs to be disposed in various landfills, consequently posing a serious threat to the environment.
