---
title: "MDSS Datathon"
author: "Arindam Baruah, Aishwarya Anil Kumar"
format: html
editor: visual
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(ggplot2)
library(visdat)
library(janitor)
library(plotly)
library(ggthemes)
library(scales)
library(RColorBrewer)
```

```{r}
df <- read_csv("data-raw/data_waste.csv")
head(df)
```

Let us clean the names of the variables

```{r}
df_waste <- clean_names(df)


df_waste[c("start_year","end_year")] <- (str_split_fixed(df_waste$year,"-",2))
df_waste$start_year <- as.numeric(df_waste$start_year)
df_waste$end_year <- as.numeric(df_waste$end_year)

df_waste<- df_waste%>% select(-year) 

head(df_waste)
```

# Hazardous waste

```{r}
df_haz <- df_waste %>% filter(category == "Hazardous wastes")
df_haz_tot <- df_haz %>% filter(stream == "Total")
df_haz_total <- df_haz_tot %>% group_by(start_year,jurisdiction,category) %>% 
  summarise(Tonnes = sum(tonnes)) %>% 
  ungroup() 



```

```{r}
pl1 <- ggplot(data = df_haz_total,aes(x = start_year,y = Tonnes)) +
  geom_line() + geom_point(color='red') + theme(axis.text.x = element_text(angle = 45,size=5)) + facet_wrap(~jurisdiction) + labs(x = "Year",y = "Tons",title = "Hazardous waste generated in each state",subtitle = "2006-2021") + theme_classic()
pl1
```

1.  NSW accounts for the highest hazardous waste in the whole of Australia with close to 8.5 MT. The hazardous wastes are on a rising trend.
2.  QLD is observed to record constant generation of hazardous waste each year with approximate 6 MT.
3.  VIC is observed to record a rising trend of hazardous waste in recent years with close to 5.75 MT in 2021.

## How is the Hazardous waste dealt with in each state ?

```{r}

haz_jurisdictions <- c("NSW","Vic","Qld")
df_haz_subset <- df_haz_tot %>% filter(jurisdiction %in% haz_jurisdictions)

df_haz_man <- df_haz_subset %>% group_by(jurisdiction,management) %>% 
  summarise(Tonnes = sum(tonnes)) %>% ungroup()


pl2 <- ggplot(data = df_haz_man,aes(x = reorder(management,Tonnes),y = Tonnes/10^6)) + geom_col(fill = "lightgray",color = "black") + 
  facet_wrap(~jurisdiction) + coord_flip() + geom_col(data = ~filter(.x, management == "Landfill"),fill = "darkred") + geom_label(data = ~filter(.x, management == "Landfill"),aes(label = round(Tonnes/10^6,1),hjust=1)) + geom_col(data = ~filter(.x, management == "Energy from waste facility"),fill = "green") + geom_label(data = ~filter(.x, management == "Landfill"),aes(label = round(Tonnes/10^6,1),hjust=1)) + geom_label(data = ~filter(.x, management == "Energy from waste facility"),aes(label = round(Tonnes/10^6,1),hjust=-0.5)) + 
  labs(x = "Waste management technique",y = "Mega tonnes of waste",title ="Hazardous waste management in the top 3 \n waste producing states",subtitle = "2006-2021") + theme_economist() 
pl2
```

We observe that the:

1.  In all the three states (NSW,Qld,Vic), majority of the hazardous waste are managed through disposing in landfills. While this may be a temporary solution, the implications can be quite devastating **as such waste normally take years to decompose and start polluting the water ways as well as pottable water resources.**

2.  While there have been efforts to utilise hazardous waste to produce energy, it is still a long way behind, with more concerted efforts required by each state to curb disposal of hazardous waste into landfills and obtain clean energy.

3.  While **VIC is observed to generate lesser hazardous waste than NSW and QLD, they're also leading with the highest energy production from waste facilities** with approximately 1.6 Mega tonnes of waste utilised.

# What is the constitution of these Hazardous wastes ?

```{r}

hazardous_waste_summary <- df_haz %>%
  group_by(type) %>%
  filter(jurisdiction %in% haz_jurisdictions) %>%
  summarise(Total_Tonnage = sum(tonnes))

hazardous_waste_summary <- hazardous_waste_summary %>%
  filter(!is.na(type) & !is.na(Total_Tonnage))


ggplot(hazardous_waste_summary, aes(x = reorder(type, Total_Tonnage), y = Total_Tonnage)) +

  # Lollipop stems
  geom_segment(aes(y = 0, 
                   yend = Total_Tonnage, 
                   x = reorder(type, Total_Tonnage),
                   xend = reorder(type, Total_Tonnage)), 
               color = "black", size = 1) +
  
  # Lollipop heads
  geom_point(aes(fill = type), size = 6, shape = 21, color = "white") +

  # Discrete color palette for lollipop heads
  scale_fill_brewer(palette="Set3") +
  
  # Y-axis scale formatting
  scale_y_continuous(labels = scales::label_comma()) +
  
  # Chart title and axes labels
  labs(title = "Composition of Hazardous Wastes", 
       x = "Type of Waste", 
       y = "Total Tonnage") +

  # Classic theme and customization
  theme_classic() +
  coord_flip() +
  theme(legend.position = "none", 
        axis.text.y = element_text(size = 8, color = "black"),
        axis.title = element_text(face = "bold", size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        panel.background = element_rect(fill = "white"),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(linewidth = 0.1, color = "yellow"))


```

